---
title: "Situation Report"
author: "YOUR NAME"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: united
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this activity, you will work in groups to create a situation report (SitRep) addressing the following research question:

**How did COVID-19 mortality rates differ along the Social Vulnerability Index (SVI) spectrum in 2020?**

### Learning Objectives

- Gather: Import data from multiple source types (.csv, API)
- Tidy: Explore, wrangle, clean, transform, visualize data
- Analyze: Combine more than one dataset to address research question
- Communicate: Use figures, tables, and text

## Setup
It can be helpful to start an R project for an analysis of this scope. Here are [posit's instructions](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects) to do so. This ensures your parent directory for this SitRep contains all the necessary data and output and works seamlessly with the `here()` function. 

### Notes about the YAML header: 
This header will automatically udpate the date of the SitRep to the date that you "knit" it, i.e., publish from your source .Rmd to the .html output. 

The design of this allows for an interactive html output with multi-level headings automatically creating a table of contents (toc=TRUE). It also has the code showing by default, but allows the end user to have the option to hide code chunks (code_folding=show).

## Load packages

Common useful packages for data analysis include:

- `tidyverse` (includes dplyr, ggplot2, readr, tidyr, etc.)
- `lubridate` (date manipulation, useful to convert dates to CDC Epiweeks)

Useful functions: 

- `install.packages()`
- `library()`

```{r load_packages, message=FALSE}
# notice the R code chunk header ```{r ...}
# tip: give these unique descriptive names (helpful for debugging)
# also can provide chunk-specific preferences
# e.g., not printing package loading messages in your output with message=FALSE


# uncomment and replace with package of interest that is not already installed
# install.packages("tidyverse")

# otherwise load pre-installed packages
library(tidyverse)
library(here) # sets your working directory to your parent .Rproj or .git containing folder

# load additional packages...

```

## Load data 

You will work with multiple datasets:

- [COVID-19 cases and deaths](https://data.cdc.gov/dataset/Weekly-United-States-COVID-19-Cases-and-Deaths-by-/yviw-z6j5/about_data)
- [Social Vulnerability Index (SVI)](https://www.atsdr.cdc.gov/place-health/php/svi/svi-data-documentation-download.html)
- [County population estimates](https://www.census.gov/data/tables/time-series/demo/popest/2020s-counties-total.html)

Can download directly or use API (note API default download limits)

Useful functions:
- `read_csv()`

```{r load_data}

# load COVID-19 case and death data
# e.g., read_csv(here("data/YOURFILENAME.csv"))

# load SVI data

# load County population data


```

## Inspect and Understand Data

- Examine structure of datasets (wide/long)
- Identify key variables
- What does each row represent?

Useful functions:

- `glimpse()` from `dplyr`
- `head()` from base R
- `summary()`
- `nrow()`
- `ncol()`
- `distinct()`
- `View()`
- `table()`
- `histogram()`


```{r}

```

### Data cleaning

- Check for missingness, variable storage type
- Remove extraneous variables, or observations outside of scope of analysis

Useful functions:

- `select()`
- `rename()`
- `mutate()`
- `case_when()`
- `if_else()`
- `filter()`
- `drop_na()`
- `replace_na()`
- `separate()`
- `unite()`

```{r}

```

## Merging

- Join data sets together
- What kind of join do you think you should use?
- What are the joining variables (common to all datasets?)
- How do you check the merge worked properly? 
Hint: Does your output have the expected dimensions? Are any new NAs introduced?


Useful functions:

- `left_join()`
- `inner_join()`
- `right_join()`
- `full_join()`

```{r}


```


## Create derived variables

- Need to create a variable for COVID-19 mortality rates
- What denominator do you need for your rates?

Useful functions:
- `rename()`
- `mutate()`
- `case_when()`
- `if_else()`

## Categorize SVI 

Check the [2020 SVI data dictionary](https://svi.cdc.gov/map25/data/docs/SVI2020Documentation_08.05.22.pdf).
Which variables from the social vulnerability index make most sense to include in the analysis? Do you want to make your own composite variable?

```{r}


```

## Exploratory Analysis

- Summary statistics for key variables
- Plot distributions

Useful functions:

- `mean()`
- `sd()`
- `median()`
- `IQR()`
- `histogram()`
- `ggplot()` + `geom_density()`
- `summarize()`

```{r}


```

## Create visualizations

- Think about your audience
- What are your key variables? Continuous? Categorical?

Check out the [ggplot cheat sheet](https://rstudio.github.io/cheatsheets/data-visualization.pdf) for inspiration.

Useful functions:

Core ggplot2:

- `ggplot()`
- `aes()`

Geoms:

- `geom_point()`
- `geom_line()`
- `geom_bar()`
- `geom_col()`
- `geom_boxplot()`

Scales + labels:

- `labs()`
- `scale_y_continuous()`
- `theme_minimal()`

Faceting:

- `facet_wrap()`

```{r}

```

## Create summary table

Useful functions
- `group_by()`
- `summarize()`
- `arrange`
- `pivot_wider()`
- `pivot_longer()`

Other useful packages & functions include `tbl_summary()` from the `gtsummary` package. 

```{r}

```

## Write key findings

- Summarize main findings (bullet points)
- Interpretation 
- Limitations

