---
title: "Statewide ILI Surveillance Report"
output: pdf_document
params:
  surveillance_data: NULL
  priority_table: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)

# data required for report
surveillance_data <- params$surveillance_data
priority_table  <- params$priority_table

#instead of using params we can pull out needed information from the data supplied
current_week <- max(surveillance_data$week)
```

# Influenza-Like Illness Surveillance Report By County and Epi Week

**Report generated:** `r format(Sys.Date(), '%B %d, %Y')` up to Epi week `r current_week`

## Summary

```{r}
knitr::kable(
  priority_table %>%
    select(
      County                   = county,
      Population               = population,
      `Latest Week Cases`      = latest_cases,
      `Case Rate per 100k`     = case_rate_per_100k,
      `Z-score`                = z_score,
      Priority                 = priority
    ),
  align = "c"
) 
```
Counties report weekly case data for prior Epi week to state PHD by noon Wednesday.
$\\$
$\\$

```{r fig.width=6, fig.height=3.5}

priority_colors <- c(
  "High - deploy resources now"  = "#d62728",
  "Medium - increase monitoring" = "#ff7f0e",
  "Low - routine surveillance"   = "#2ca02c",
  "Data issue - needs follow-up" = "#7f7f7f"
)


# Process and plot the data
priority_table %>%
  # 1. Create a temporary numeric ordering column where NA becomes -1
  mutate(order_helper = replace_na(case_rate_per_100k, -1)) %>%
  # 2. Reorder the county factor based on our bulletproof helper column
  mutate(county = fct_reorder(county, order_helper)) %>%
  # 3. Replace NA with the maximum value so Elkridge fills the full bar width
  mutate(case_rate_per_100k = replace_na(case_rate_per_100k, max(case_rate_per_100k, na.rm = TRUE))) %>%
  # 4. Generate the text string ONLY for the data issue row
  mutate(bar_label = if_else(priority == "Data issue - needs follow-up", 
                             "Data Error, Follow Up Required", 
                             "")) %>%
  # 5. Build the ggplot
  ggplot(aes(x = county, y = case_rate_per_100k, fill = priority)) +
  geom_col() +
  # Add bold text inside the bar, slightly shifted right from the axis boundary
  geom_text(aes(label = bar_label), y = 5, hjust = 0, color = "white", fontface = "bold") +
  scale_fill_manual(values = priority_colors) +
  coord_flip() +
  labs(
    title = "ILI Case Rate per 100,000 Residents by County",
    subtitle = paste("For most current data, epi week:", current_week ),
    x = NULL,
    y = "Cases per 100,000",
    fill = "Resource Priority"
  ) +
  theme_minimal()
```


## Weekly Case Trend

```{r fig.width=6, fig.height=3.5}

# 1. Standardize the cases into a rate per 100,000 people
surveillance_data %>%
  mutate(case_rate_per_100k = (cases / population) * 100000) %>%
  # 2. Initialize the ggplot
  ggplot(aes(x = week, y = case_rate_per_100k, color = county, group = county)) +
  # Add lines and small points at each week mark
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  # 3. Clean up the design
  scale_x_continuous(breaks = 1:12) +  # Forces whole numbers on the week axis
  labs(
    title = "Weekly ILI Case Rate Trends by County",
    subtitle = "Standardized comparison across epi weeks",
    x = "Epi Week",
    y = "Cases per 100,000 Residents",
    color = "County"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(), # Removes unnecessary grid lines
    legend.position = "right"
  )

```


```{r fig.width=6, fig.height=3.5}

# 1. Initialize the ggplot using raw cases
surveillance_data %>%
  ggplot(aes(x = week, y = cases, color = county, group = county)) +
  
  # Add lines and small points at each week mark
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  
  # 2. Clean up the axes and design
  scale_x_continuous(breaks = min(surveillance_data$week):max(surveillance_data$week)) +
  labs(
    title = "Weekly ILI Absolute Case Trends by County",
    subtitle = "Raw case counts by epi week",
    x = "Epi Week",
    y = "Number of Cases",
    color = "County"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(), # Removes unnecessary background grid lines
    legend.position = "right"
  )

```



---

*This report was generated automatically from state ILI surveillance data. For questions, contact the state epidemiology office.*
